[Re] How learning can guide evolution

### 2017-08-20 article/stojic2017.tex
corrected grammar in the article, labels in the figures closer to lines and removed axis

  {\bfseries Received}  Aug, 17, 2017\\
  {\bfseries Accepted}  Sep, 1, 2015\\
  {\bfseries Published} Sep, 1, 2015\\

@@

Lamarckian hypothesis that experiences accumulated during the lifetime
of an individual organism are directly transmitted to the next
generation is considered to be incorrect. The reference paper
\autocite{hinton1987learning} argues that this does not necessarily mean
that individual learning cannot exert influence on evolution in an
indirect fashion, providing a simple but convincing computational
example of such interaction. The authors use an extreme scenario where
optimization landscape is completely flat except for a single peak. In
such environment curvature of the optimization surface does not provide
any guidance and evolutionary search alone cannot find the maximum.
Through a simulation the authors show that such hostile environment can
be tackled by a combination of evolutionary and individual learning. If
organisms can search during their lifetime in addition, those organisms
whose genomes are closer to the targeted peak are going to be able to
find the peak with individual search. These organisms will have higher
fitness and transmit their genes to the next generation. In effect,
capacity for individual learning ``creates a hill'' to the peak that the
evolution can climb, as illustrated in Figure 1 in
\textcite{hinton1987learning}\footnote{Note that captions of the figures
  are switched, caption of Figure 1 is incorrectly placed under Figure
  2, and vice versa}. This is one of the clearest illustrations of
benefits of individual learning for evolution and interaction between
the two learning processes.\footnote{An indirect influence of individual
  learning on evolutionary learning is often called Baldwin effect.}
This result has made a substantial impact in cognitive science
literature.

The main aim of this article is to replicate the simulations reported in
the reference paper \autocite{hinton1987learning}. I have inquired with
one of the authors and the original implementation of the simulations is
not available any more. I am also not aware of other implementations
elsewhere. Simulations are relatively simple and I propose a replication
based on description from the reference paper. Replication code relies
on R programming language \autocite{R} and several R packages
\autocites{ggplot2}{dplyr}{reshape2}{doParallel}{foreach}{doRNG}.

@@

Description of simulations can be found in the caption of Figure 1 in
the reference paper \autocite{hinton1987learning}. I consider the
description to be detailed enough for an implementation. I have followed
it closely, with two exceptions. First, I have repeated evolution many
times to average out potential noise in the process. In the reference
article such iterations are not explicitly mentioned, and moreover, when
interpreting their results from Figure 2, the authors state that Figure
2 presents results of ``\ldots{} a typical evolutionary search
\ldots{}'', suggesting it is a single run of the simulation. Second, I
let the population of agents evolve for 1000 generations instead of 50
generations reported in \textcite{hinton1987learning}. Reason for this
will become clear in the Results section.

@@

\hypertarget{tbl:parameters}{}
\begin{longtable}[]{@{}lll@{}}
\caption{\label{tbl:parameters}Simulation parameters. }\tabularnewline
\toprule
Table & Value & Description\tabularnewline
\midrule
\endfirsthead
\toprule
Table & Value & Description\tabularnewline
\midrule
\endhead
noSim & 100 & Number of times evolution is simulated\tabularnewline
noGenerations & 1000 & Number of generations for evolutionary
algorithm\tabularnewline
noAgents & 1000 & Number of agents in a population\tabularnewline
lifetime & 1000 & Number of cycles available for individual
learning\tabularnewline
noAlleles & 20 & Number of alleles in agent's genome\tabularnewline
\(p_0\) & 0.25 & Probability that allele is of type zero\tabularnewline
\(p_1\) & 0.25 & Probability that allele is of type one\tabularnewline
\(p_?\) & 0.50 & Probability that allele is of type ?\tabularnewline
\bottomrule
\end{longtable}

@@

Simulation results are summarized in Figure ~\ref{fig:relFrequencies50},
which corresponds to Figure 2 from the reference article. Qualitatively
the results are very similar - there is an obvious increase in
proportion of alleles correctly set by the evolution (close to 75\% by
generation 50) and proportion of incorrect alleles decreases to zero by
generation 50. This is a clear evidence of individual learning guiding
evolutionary search even though there is no direct transmission of
knowledge acquired during individual learning from one generation to the
next.

There is an important difference with respect to the reference paper,
however. In original simulations proportion of Undecided alleles that
are left to individual learning is relatively constant, close to initial
50\%. The authors point out this as an interesting result - there is
little selective pressure to specify last few connections by evolution
because few switches are quickly set through individual learning. In
contrast, results in Figure ~\ref{fig:relFrequencies50} show that
proportion of Undecided alleles is continuously decreasing. In fact,
whereas proportion of Correct alleles in the reference article flattens
out in my case it continues increasing, with increase coming from
proportion of Undecided alleles decreasing further. To make sure this is
really a long run trend I have let populations evolve for larger number
of generations than in the reference article. Results of all 1000
generations are presented in Figure ~\ref{fig:relFrequencies1000}. By
generation 1000 proportion of Undecided alleles drops to 9\%. Even
though decrease is smaller with each generation, it does not seem to
stop and flatten out by the end of the simulation.

What is the source of this particular difference in results? One
explanation is that results reported in Figure 2 in the reference
article stem from a non-typical draw. This is unlikely as upon examining
evolutionary paths of Undecided alleles in all simulation runs, they all
show a strong downward trend already in first 50 generations, differing
only in the exact generation when this decrease onsets. Another
explanation is that original implementation is flawed as there actually
is a selection pressure for eliminating individual learning after some
time. The fitness function described in the reference article is such
that the smaller the number of steps an agent takes in individual
lifetime to reach the peak, the higher the probability of mating and
transmitting the genes to the next generation. Number of steps is a
direct function of number of undecided alleles in the genome. Hence,
there is a selection pressure for decreasing the number of Undecided
alleles. In personal correspondence, one of the authors agreed with this
interpretation, stating that indeed there should be a selective pressure
for decrease in number of Undecided alleles. The pressure should
decrease as a power function of the mean number of steps agents need to
reach the peak, which explains rapidly decreasing improvements with each
generation.

\begin{figure}
\centering
\includegraphics{Figure2.pdf}
\caption{The evolution of the relative frequencies of the three possible
types of allele. Correct alleles are those that are set by evolutionary
learning and correspond to the targeted pattern, Incorrect denote
proportion of incorrectly set alleles, while Undecided denote proportion
of alleles left for individual learning. Proportions are means of 100
simulation runs, and barely visible grey ribbons are standard errors.
This is the reproduction of Figure 2 in the original
article.}\label{fig:relFrequencies50}
\end{figure}

@@

Qualitatively the result is comparable to that reported in the reference
article, but there is relatively important quantitative difference.
Proportions of Undecided alleles in implementation proposed here
diminish over time, whereas in the original implementation they stay
close to initial values. I attribute the difference to a flaw in the
original implementation, providing arguments based on fitness function
proposed in the reference article. Although the main result of the
reference article - individual learning indirectly guiding evolutionary
search - has been reproduced, the difference I have found suggests that
individual learning would be eradicated over time.

>>>>

  {\bfseries Received}  Aug, 20, 2017\\
  {\bfseries Accepted}  Sep, 1, 2015\\
  {\bfseries Published} Sep, 1, 2015\\

@@

The Lamarckian hypothesis that adaptations accumulated during the
lifetime of an individual organism are directly transmitted to the next
generation is considered to be incorrect. The reference article
\autocite{hinton1987learning} argues that this does not necessarily mean
that individual learning cannot exert influence on evolution in an
indirect fashion, providing a simple but convincing computational
example of such an interaction. The authors use an extreme scenario
where the optimization landscape is completely flat except for a single
peak. In such an environment, curvature of the optimization surface does
not provide any guidance and evolutionary search alone cannot find the
maximum. Through a simulation the authors show that such a hostile
environment can be tackled by a combination of evolutionary and
individual learning. In addition, if organisms can search during their
lifetime, those organisms whose genomes are closer to the targeted peak
are going to be able to find the peak with individual search. Such
organisms will have higher fitness and transmit their genes to the next
generation. In effect, capacity for individual learning ``creates a
hill'' to the peak that evolutionary search can climb, as illustrated in
Figure 1 in \textcite{hinton1987learning}\footnote{Note that captions of
  the figures are switched, caption of Figure 1 is incorrectly placed
  under Figure 2, and vice versa}. This is one of the clearest
illustrations of the benefits of individual learning for evolution and
the interaction between the two learning processes.\footnote{An indirect
  influence of individual learning on evolutionary learning is often
  called Baldwin effect.} This result has made a substantial impact in
the cognitive science literature.

The main aim of this article is to replicate the simulations reported in
the reference article \autocite{hinton1987learning}. I have corresponded
with one of the authors and learned that the original implementation of
the simulations is no longer available. I am also not aware of other
implementations elsewhere. The simulations are relatively simple, and I
thus propose a replication based on description from the reference
article. Replication code relies on R programming language \autocite{R}
and several R packages
\autocites{ggplot2}{dplyr}{reshape2}{doParallel}{foreach}{doRNG}{ggrepel}{directlabels}.

@@

A description of the simulations can be found in the caption of Figure 1
in the reference article \autocite{hinton1987learning}. I consider the
description to be detailed enough for the implementation. I have
followed it closely, with two exceptions. First, I have repeated the
evolutionary search many times to average out potential noise. In the
reference article, such iterations are not explicitly mentioned, and
moreover, when interpreting their results from Figure 2, the authors
state that Figure 2 presents results of ``\ldots{} a typical
evolutionary search \ldots{}'', suggesting it is a single run of the
simulation. Second, I let the population of agents evolve for 1000
generations instead of 50 generations reported in
\textcite{hinton1987learning}. Reason for this will become clear in the
Results section.

@@

\hypertarget{tbl:parameters}{}
\begin{longtable}[]{@{}lll@{}}
\caption{\label{tbl:parameters}Simulation parameters. }\tabularnewline
\toprule
Table & Value & Description\tabularnewline
\midrule
\endfirsthead
\toprule
Table & Value & Description\tabularnewline
\midrule
\endhead
noSim & 100 & Number of times evolutionary search is
repeated\tabularnewline
noGenerations & 1000 & Number of generations for evolutionary
algorithm\tabularnewline
noAgents & 1000 & Number of agents in a population\tabularnewline
lifetime & 1000 & Number of cycles available for individual
learning\tabularnewline
noAlleles & 20 & Number of alleles in agent's genome\tabularnewline
\(p_0\) & 0.25 & Probability that allele is of type zero\tabularnewline
\(p_1\) & 0.25 & Probability that allele is of type one\tabularnewline
\(p_?\) & 0.50 & Probability that allele is of type ?\tabularnewline
\bottomrule
\end{longtable}

@@

The simulation results are summarized in Figure
~\ref{fig:relFrequencies50}, which corresponds to Figure 2 from the
reference article. Qualitatively, the results are very similar - there
is an obvious increase in the proportion of alleles correctly set by the
evolutionary learning (close to 75\% by generation 50) and proportion of
incorrect alleles decreases to zero by generation 50. This is clear
evidence of individual learning guiding evolutionary search, even though
there is no direct transmission of knowledge acquired during individual
learning from one generation to the next.

\begin{figure}
\centering
\includegraphics{Figure2.pdf}
\caption{The evolution of the relative frequencies of the three possible
types of allele. Correct alleles are those that are set by evolutionary
learning and correspond to the targeted pattern. Incorrect alleles
represent the proportion of incorrectly set alleles, while Undecided
represent the proportion of alleles left for individual learning.
Proportions are means of 100 simulation runs, and barely visible grey
ribbons are standard errors. This is the reproduction of Figure 2 in the
reference article
\autocite{hinton1987learning}.}\label{fig:relFrequencies50}
\end{figure}

There is an important difference with respect to the reference article,
however. In the original simulations, the proportion of Undecided
alleles that are left to individual learning is relatively constant,
close to initial 50\%. The authors point this out as an interesting
result, suggesting that there is little selective pressure to specify
last few connections by evolutionary learning because those can be
quickly set through individual learning. In contrast, the results in
Figure ~\ref{fig:relFrequencies50} show that the proportion of Undecided
alleles is continuously decreasing. In fact, whereas the proportion of
Correct alleles in the reference article flattens out, in my case it
continues increasing, with the increase coming from the proportion of
Undecided alleles decreasing further. To make sure that this is really a
long run trend I have let populations evolve for a larger number of
generations than in the reference article. Results of all 1000
generations are presented in Figure ~\ref{fig:relFrequencies1000}. By
generation 1000 the proportion of Undecided alleles drops to 9\%. Even
though the decrease is smaller with each generation, it does not seem to
stop and flatten out by the end of the simulation.

What is the source of this particular difference in results? One
explanation is that the results reported in Figure 2 in the reference
article stem from a non-typical draw. This is unlikely, as upon
examining the evolutionary paths of Undecided alleles in all simulation
runs, they all show a strong downward trend already in the first 50
generations, differing only in the exact generation when this decrease
onsets. Another explanation is that the original implementation is
flawed as there actually is a selection pressure for eliminating
individual learning after some time. The fitness function described in
the reference article is such that the smaller the number of steps an
agent takes in an individual lifetime to reach the peak, the higher the
probability of mating and transmitting genes to the next generation. The
number of steps is a direct function of the number of undecided alleles
in the genome. Hence, there is a selection pressure for decreasing the
number of Undecided alleles. In personal correspondence, one of the
authors agreed with this interpretation, stating that indeed there
should be a selective pressure for a decrease in number of Undecided
alleles. The pressure should decrease as a power function of the mean
number of steps an agent needs to reach the peak, which explains rapidly
decreasing improvements with each generation.

@@

Qualitatively, the result is comparable to that reported in the
reference article, but there are relatively important differences. The
proportion of Undecided alleles in the implementation proposed here
diminish over time, whereas in the original implementation they stay
close to the initial values. I attribute the difference to a flaw in the
original implementation, providing arguments based on a fitness function
proposed in the reference article. Although the main result of the
reference article - that individual learning can indirectly guide
evolutionary search - has been reproduced, the difference in simulation
results suggests that individual learning might be eradicated over time.
This decrease in usefulness of individual learning occurs in static
environments where targeted genome stays fixed, as in these simulations.
In more realistic scenarios, where environments are non-stationary and
the targeted genome would be changing over time, individual learning
would have a more stable role.
